Examining spatial equity of bike share - a balanced floating catchment area approach

### 2021-03-31 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Final submission

---
title: Examining horizontal and vertical equity to a public bicycle share program using a balanced floating catchment area accessibility approach 
author:
  - name: Elise Desjardins
    email: desjae@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliation: University of Toronto Scarborough
    footnote: 
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 
address:
  - code: McMaster University
    address: School of Earth, Environment & Society, McMaster University, 1280 Main Street West, Hamilton, ON L8S4L8
  - code: University of Toronto Scarborough
    address: Department of Geography & Planning, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON M1C1A4
footnote:
  - code: 1
    text: "Corresponding Author"
abstract:

@@

journal: "Transportation Research Part D: Transport and Environment"
date: "`r Sys.Date()`"
bibliography: [packages.bib, references.bib]
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

@@

On the whole, existing studies highlight the need for PBSPs to be more highly accessible for diverse populations in order to increase use beyond the "typical" users. This has been the focus of recent research [see, among others, @auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020]. Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs. A few North American cities have launched specific programs to address barriers to bike share or to expand service to more deprived areas: Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018]. However, exploring transportation equity by investigating where bicycle share stations are located, often using neighborhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones. Meaning that, stations may be lacking in certain neighborhoods but there may be stations accessible within a reasonable walking time. This is where geographical accessibility becomes an important consideration.

@@

```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bicycle share stations in Hamilton, Ontario. The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in pink. Hamilton Census Metropolitan Area is shown in grey."}
ggplot() + 
  geom_sf(data = cropped_da,
          color = NA,
          fill = "gray") +
  geom_sf(data = sobi_service, 
          fill = NA,
          color = "black") +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          size = 1,
          color = "pink") +
  geom_sf(data = sobi_hubs %>% 
            dplyr::filter(hub_type == "Conventional" & 
                            hub_status == "Active"), # Remove deactivated hubs
          aes(size = RACKS_AMOU),
          shape = 21,
          fill = "blue") +
  geom_sf(data =  sobi_hubs %>% 
            dplyr::filter(hub_type == "ERI"), 
          aes(size = RACKS_AMOU), 
          shape = 21,
          fill = "green") +
  scale_size(name = "Number of racks", range = c(0.5, 4)) +
  theme(panel.background = element_rect(fill = "white"))
```

@@

Figures 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps that combine the spatial distribution of accessibility and median total household income, using tertiles for the coloring scheme. In each figure, the top panel is the system without the equity stations (i.e., the original system configuration), and the bottom panel is accessibility in the current system with equity stations. As expected, the analysis indicates that the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service area of stations. We find that stations added to Hamilton's public bicycle share program did indeed expand the spatial coverage of the system, leading to more balance across the service area. By implementing equity stations in areas with lower median total household income in Hamilton, the PBSP attained greater horizontal equity by extending the spatial distribution of bicycles across the city. This is particularly evident at the minimum and average thresholds of three and five minutes, respectively, where the equity stations fill a number of gaps in program coverage.

>>>>

---
title: Examining spatial equity of bike share - a balanced floating catchment area approach 
author:
  - name: Elise Desjardins
    email: desjae@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliation: University of Toronto Scarborough
    footnote: 
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 
address:
  - code: McMaster University
    address: School of Earth, Environment & Society, McMaster University, 1280 Main Street West, Hamilton, ON L8S4L8
  - code: University of Toronto Scarborough
    address: Department of Geography & Planning, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON M1C1A4
footnote:
  - code: 1
    text: "Corresponding Author"
abstract:

@@

journal: "Transportation Research Part D: Transport and Environment"
date: "`r Sys.Date()`"
bibliography: [packages.bib, references.bib]
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
header-includes:
    - \usepackage{setspace}\doublespacing
---

@@

On the whole, existing studies highlight the need for PBSPs to be more highly accessible for diverse populations in order to increase use beyond the "typical" users. This has been the focus of recent research [see, among others, @auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020]. Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs. A few North American cities have launched specific programs to address barriers to accessing bike share or to expand service to more deprived areas: Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018]. However, exploring transportation equity by investigating where bicycle share stations are located, often using neighborhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones. Meaning that, stations may be lacking in certain neighborhoods but there may be stations accessible within a reasonable walking time. This is where geographical accessibility becomes an important consideration.

@@

```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bicycle share stations in Hamilton, Ontario. Conventional stations are in blue and equity stations are in green. The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in pink. Hamilton Census Metropolitan Area is shown in grey."}
ggplot() + 
  geom_sf(data = cropped_da,
          color = NA,
          fill = "gray") +
  geom_sf(data = sobi_service, 
          fill = NA,
          color = "black") +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          size = 1,
          color = "pink") +
  geom_sf(data = sobi_hubs %>% 
            dplyr::filter(hub_type == "Conventional" & 
                            hub_status == "Active"), # Remove deactivated hubs
          aes(size = RACKS_AMOU),
          shape = 21,
          fill = "blue") +
  geom_sf(data =  sobi_hubs %>% 
            dplyr::filter(hub_type == "ERI"), 
          aes(size = RACKS_AMOU), 
          shape = 21,
          fill = "green") +
  scale_size(name = "Number of racks", range = c(0.5, 4)) +
  theme(panel.background = element_rect(fill = "white"))
```

@@

Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} depict bivariate choropleth maps that combine the spatial distribution of accessibility and median total household income, using tertiles for the coloring scheme. In each figure, the top panel is the system without the equity stations (i.e., the original system configuration), and the bottom panel is accessibility in the current system with equity stations. As expected, the analysis indicates that the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service area of stations. We find that stations added to Hamilton's public bicycle share program did indeed expand the spatial coverage of the system, leading to more balance across the service area. By implementing equity stations in areas with lower median total household income in Hamilton, the PBSP attained greater horizontal equity by extending the spatial distribution of bicycles across the city. This is particularly evident at the minimum and average thresholds of three and five minutes, respectively, where the equity stations fill a number of gaps in program coverage.

### 2021-03-31 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Final submission

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining horizontal and vertical equity to a public bicycle share program using a balanced floating catchment area accessibility approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

  \title{Examining horizontal and vertical equity to a public bicycle share
program using a balanced floating catchment area accessibility approach}
    \author[McMaster University]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
    \author[McMaster University]{Antonio Paez}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, McMaster University, 1280 Main
Street West, Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
1265 Military Trail, Toronto, ON M1C1A4}
      \cortext[1]{Corresponding Author}

@@

On the whole, existing studies highlight the need for PBSPs to be more
highly accessible for diverse populations in order to increase use
beyond the ``typical'' users. This has been the focus of recent research
(see, among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
MacArthur et al., 2020). Offering more people the option of using
sustainable and active transportation, particularly those who have lower
socioeconomic status and might benefit the most, is a worthy policy goal
for cities with PBSPs. A few North American cities have launched
specific programs to address barriers to bike share or to expand service
to more deprived areas: Philadelphia, Pennsylvania (Caspi and Noland,
2019), and Hamilton, Ontario (Hosford and Winters, 2018). However,
exploring transportation equity by investigating where bicycle share
stations are located, often using neighborhoods or census tracts as the
geographical unit of analysis, can ignore or miss the benefits that may
be derived from adjacent zones. Meaning that, stations may be lacking in
certain neighborhoods but there may be stations accessible within a
reasonable walking time. This is where geographical accessibility
becomes an important consideration.

@@

\caption{The spatial distribution of bicycle share stations in Hamilton, Ontario. The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in pink. Hamilton Census Metropolitan Area is shown in grey.}\label{fig:sobi-stations-in-hamilton}
\end{figure}

@@

Figures 1 {[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
{[}\ref{fig:figure-bi-map-threshold-15}{]} depict bivariate choropleth
maps that combine the spatial distribution of accessibility and median
total household income, using tertiles for the coloring scheme. In each
figure, the top panel is the system without the equity stations (i.e.,
the original system configuration), and the bottom panel is
accessibility in the current system with equity stations. As expected,
the analysis indicates that the extreme threshold of fifteen minutes is
associated with the largest number of people who are within the assumed
service area of stations. We find that stations added to Hamilton's
public bicycle share program did indeed expand the spatial coverage of
the system, leading to more balance across the service area. By
implementing equity stations in areas with lower median total household
income in Hamilton, the PBSP attained greater horizontal equity by
extending the spatial distribution of bicycles across the city. This is
particularly evident at the minimum and average thresholds of three and
five minutes, respectively, where the equity stations fill a number of
gaps in program coverage.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining spatial equity of bike share - a balanced floating catchment area approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{setspace}\doublespacing
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

  \title{Examining spatial equity of bike share - a balanced floating catchment
area approach}
    \author[McMaster University]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
    \author[McMaster University]{Antonio Paez}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, McMaster University, 1280 Main
Street West, Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
1265 Military Trail, Toronto, ON M1C1A4}
      \cortext[1]{Corresponding Author}

@@

On the whole, existing studies highlight the need for PBSPs to be more
highly accessible for diverse populations in order to increase use
beyond the ``typical'' users. This has been the focus of recent research
(see, among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
MacArthur et al., 2020). Offering more people the option of using
sustainable and active transportation, particularly those who have lower
socioeconomic status and might benefit the most, is a worthy policy goal
for cities with PBSPs. A few North American cities have launched
specific programs to address barriers to accessing bike share or to
expand service to more deprived areas: Philadelphia, Pennsylvania (Caspi
and Noland, 2019), and Hamilton, Ontario (Hosford and Winters, 2018).
However, exploring transportation equity by investigating where bicycle
share stations are located, often using neighborhoods or census tracts
as the geographical unit of analysis, can ignore or miss the benefits
that may be derived from adjacent zones. Meaning that, stations may be
lacking in certain neighborhoods but there may be stations accessible
within a reasonable walking time. This is where geographical
accessibility becomes an important consideration.

@@

\caption{The spatial distribution of bicycle share stations in Hamilton, Ontario. Conventional stations are in blue and equity stations are in green. The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in pink. Hamilton Census Metropolitan Area is shown in grey.}\label{fig:sobi-stations-in-hamilton}
\end{figure}

@@

Figures \ref{fig:figure-bi-map-threshold-3},
\ref{fig:figure-bi-map-threshold-5},
\ref{fig:figure-bi-map-threshold-10}, and
\ref{fig:figure-bi-map-threshold-15} depict bivariate choropleth maps
that combine the spatial distribution of accessibility and median total
household income, using tertiles for the coloring scheme. In each
figure, the top panel is the system without the equity stations (i.e.,
the original system configuration), and the bottom panel is
accessibility in the current system with equity stations. As expected,
the analysis indicates that the extreme threshold of fifteen minutes is
associated with the largest number of people who are within the assumed
service area of stations. We find that stations added to Hamilton's
public bicycle share program did indeed expand the spatial coverage of
the system, leading to more balance across the service area. By
implementing equity stations in areas with lower median total household
income in Hamilton, the PBSP attained greater horizontal equity by
extending the spatial distribution of bicycles across the city. This is
particularly evident at the minimum and average thresholds of three and
five minutes, respectively, where the equity stations fill a number of
gaps in program coverage.
